\huge{Bayesian RKHS-based methods in functional regression}

### 2022-09-19 slides/defense.tex
Finish defense slides

@@

\definecolor{Maroon}{cmyk}{0, 0.87, 0.88, 0.1}
\definecolor{teal}{rgb}{0.0, 0.45, 0.45}

\usetheme[block=fill, subsectionpage=progressbar, titleformat section=smallcaps]{metropolis}
\setbeamertemplate{frametitle continuation}[roman]
\setbeamerfont{section title}{size=\huge}
\setbeamertemplate{section in toc}[sections numbered]
%\setbeamertemplate{subsection in toc}[subsections unnumbered]
%\setsansfont[BviejoFont={Fira Sans SemiBold}]{Fira Sans Book}  % Increase font weigth
\widowpenalties 1 10000
\raggedbottom

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

% COLORS
\setbeamercolor{palette primary}{bg=teal}
\setbeamercolor{progress bar}{use=Maroon, fg=Maroon}

\usepackage{fontspec}
\usepackage[autostyle=true]{csquotes}
\usepackage[absolute,overlay]{textpos}
\usepackage[english]{babel}
\usepackage{microtype}
\usepackage{subcaption}
\usepackage{epigraph}
\usepackage[export]{adjustbox}
\usepackage{hyperref}
\usepackage{bm}
\usepackage{amssymb, amsmath, amsthm, amsfonts, amscd, dsfont}
\usepackage{listings}

@@

@@

% FONTS
%\usefonttheme{professionalfonts}
%\usepackage{mathpazo}
%\usepackage{eulervm}

\definecolor{backg}{HTML}{F2F2F2} % Fondo
\definecolor{comments}{HTML}{a8a8a8} % Comentarios
\definecolor{keywords}{HTML}{08388c} % Palabras clave
\definecolor{strings}{HTML}{0489B1}  % Strings

@@

\renewcommand{\baselinestretch}{1.3}
\definecolor{mLightBrown}{HTML}{f97e0b}
% \definecolor{mLightBrown}{HTML}{990000}  % Darker accent color
\newcommand\maroon[1]{\color{mLightBrown}#1\color{mDarkTeal}}

\renewcommand{\epsilon}{\varepsilon}

@@

% DOCUMENTO

@@

\begin{frame}{Functional linear regression}
  \textbf{First problem}: linear regression with functional data.
  \vspace{2em}

@@

\begin{frame}{Functional logistic regression}
  \textbf{Second problem}: logistic regression with functional data.
\vspace{2em}

@@

\begin{frame}{Shortcomings of the \(\bm{L^2}\)-models}
  \begin{itemize}
    \item Direct optimization is hopeless: \(L^2[0,1]\) is a \maroon{extremely wide space}.
    \item The usual least-squares theory cannot be applied directly; some regularization or dimensionality reduction techniques are required.
    \item They do not include simple models based on linear combinations of the marginal of the process as particular cases, such as
    \[
      Y = \alpha_0 + \sum_{j=1}^p \beta_j X(t_j) + \varepsilon.
    \]
    \item In the logistic case, the maximum likelihood estimator does not exist with probability one under fairly general conditions.
  \end{itemize}
\end{frame}

@@

\vspace{1em}
  We consider a functional parameter \maroon{\(\alpha \in \mathcal H(K)\)}, where \(\mathcal H(K)\) is the RKHS associated with the covariance function \(K(t, s)=\mathbb E[X(t)X(s)]\) of the centered process \(X\). Moreover, we substitute \(\dotprod{X}{\beta}_2\) for \maroon{\(\dotprod{X}{\alpha}_K\)}.
\end{frame}

@@

  \begin{block}{Proposition}
    \begin{enumerate}
      \item All the evaluation operators \(\delta_t(f)=f(t)\) are continuous \maroon{(characterization)}.
      \item Norm convergence implies pointwise convergence.
      \item If \(K\) is continuous, then so is every function \(f\in\mathcal H(K)\).
    \end{enumerate}
  \end{block}
\end{frame}

% \begin{frame}{RKHS's in a nutshell (cont.)}
%   There are other equivalent definitions of \(\Hcal(K)\), mainly based on the \textit{covariance operator}
% \[
% \mathcal Kf(\cdot) = \int_0^1 K(s, \cdot)f(s)\, ds, \quad f\in L^2[0, 1].
% \]
%
% \begin{definition}
% \begin{enumerate}
%   \item  \(\Hcal(K) = \mathcal K^{1/2}(L^2[0, 1])\), with \(\dotprod{f}{g}_K = \dotprod{\mathcal K^{-1/2}(f)}{\mathcal K^{-1/2}(g)}_2\).
%   \item \(\Hcal(K) = \{f \in L^2[0, 1]: \ \sum_j \lambda_j^{-1}\dotprod{f}{\phi_j}^2 < \infty \}\), with \(\dotprod{f}{g}_K = \sum_j \lambda_j^{-1}\dotprod{f}{\phi_j}\dotprod{g}{\phi_j}_2\). Here, \(\lambda_j\to 0\) and \(\phi_j\) are the eigenvalues and (orthonormal) eigenfunctions of \(\mathcal K\).
% \end{enumerate}
% \end{definition}
%
% The second definition emphasizes the fact that functions in \(\Hcal(K)\) are ``simple'', in the sense that their components in an orthonormal basis need to vanish quickly.
%
% \end{frame}

@@

  Generally speaking, the concrete realizations \(x=X(\omega)\) \maroon{do not belong to \(\Hcal(K)\) with probability \(1\)}, so the expression \(\dotprod{x}{\alpha}_K\) lacks meaning. However, we can make use of the relationship between \(\Hcal(K)\) and \(\mathcal L(X)\), the linear span of the process \(X\) within \(L^2(\Omega)\).

  \vspace{1em}
  \begin{block}{Loève's isometry}
    It is the continuous extension of the correspondence
    \[
\sum_{j=1}^p a_j X(t_j) \longleftrightarrow \sum_{j=1}^p a_j K(t_j, \cdot).
    \]
    Formally, \(\Psi^{-1}_X(U)(t) := \E[U X(t)]\) for \(U \in \mathcal L(X)\).
  \end{block}
  With a slight abuse of notation, we will write \maroon{\(\dotprod{x}{\alpha}_K \equiv [\Psi_X(\alpha)](\omega) := \Psi_x(\alpha)\)}\ for \(x=X(\omega)\) and \(\alpha \in \Hcal(K)\).
\end{frame}

@@

\begin{frame}{\ldots and two new models}
  Given a data set \(\mathcal D_n = \{(X_i, Y_i): i=1,\dots, n\}\) composed of i.i.d. samples of \((X, Y)\), for \(\alpha \in \Hcal_0(K)\) we consider the following models:

@@

      \begin{block}{Bayes's formula}
      \[
      \pi(\theta \mid \mathcal D_n) \propto \left( \prod_{i=1}^n \pi(Y_i\mid X_i, \theta) \right)\pi(\theta).
      \]
  \end{block}

@@

\begin{frame}{Prior distributions}
  Same prior scenario for both linear and logistic models.
  \vspace{.5em}
  \begin{block}{Prior for the parameters involved}
    \vspace{-1em}
  \begin{align*}
  \pi(\alpha_0, \sigma^2)              & \propto 1/\sigma^2,                                                     \\
  \tau                     & \sim \mathcal U([0, 1]^p),                                              \\
  b\mid \tau, \sigma^2 & \sim \mathcal N_p\big(b_0, g\sigma^2{\underbrace{\left[\mathcal X_\tau' \mathcal X_\tau + \eta I\right]}_{G_\tau}}^{-1}\big),
\end{align*}

\vspace{-1em}
where \(I\) is the identity matrix, \(\mathcal X_\tau = (X_i(t_j))_{i,j}\) is the data matrix, and \(b_0\in \R^p, \ g \in \R\) and \(\eta \in \R^+\) are hyperparameters.
\end{block}

@@

\begin{frame}{Posterior distributions}
  \begin{block}{Linear log-posterior}
    \vspace{-1em}
    \begin{align*}
  \log \pi(\theta\mid \mathcal D_n) & \propto \frac{1}{2}\log |G_\tau| - (p+n+2)\log \sigma\\
  &-\frac{1}{2\sigma^2} \left(\|\bm{Y}-\alpha_0\bm{1} - \mathcal X_\tau b\|^2 + \frac{1}{g}(b - b_0)'G_\tau(b - b_0) \right),
\end{align*}
where \(\bm Y=(Y_1,\dots,Y_n)'\) and \(\bm{1}\) is an \(n\)-dimensional vector of ones.
  \end{block}
  \vspace{1em}
  \begin{block}{Logistic log-posterior}
      \vspace{-1em}
    \begin{align*}
  \log \pi(\theta \mid \mathcal D_n) & \propto \sum_{i=1}^n \left[ \left(\alpha_0 + \dotprod{X_i}{\alpha}_K\right)Y_i - \log\left(1 + \exp\left\{\alpha_0 + \dotprod{X_i}{\alpha}_K\right\}\right)\right]\\
  \quad &+ \frac{1}{2}\log |G_\tau| - (p+2)\log \sigma -\frac{1}{2g\sigma^2} (b - b_0)'G_\tau(b - b_0).
\end{align*}
Remember that \(\dotprod{X_i}{\alpha}_K = \sum_j \beta_j X_i(t_j)\).
  \end{block}
\end{frame}

@@

  \begin{figure}
    \centering
    \includegraphics[width=.85\textwidth]{posterior}
  \end{figure}

@@

\begin{frame}{Predictions}
  \textbf{Summarize-then-predict}. A summary of the posterior distribution \(\theta| \mathcal D_n\) is obtained via a point-estimate statistic (mean, median, mode), and then predictions are made according to the assumed model, e.g.:
  \[
  \hat Y_i =\hat \alpha_0 + \sum_{j=1}^p \hat \beta_j X_i(\hat t_j), \quad i=1,\dots, n.
  \]

@@

\begin{frame}{Predictions (cont.)}
    We are essentially approximating the \textit{posterior predictive distribution}
\[
\pi(y\mid x, \mathcal D_n) = \int_{\Theta} \pi(y\mid x, \theta)\pi(\theta\mid \mathcal D_n)\, d\theta.
\]
  \begin{figure}
    \includegraphics[width=.675\textwidth]{ppc_linear}
  \end{figure}
\end{frame}

\begin{frame}{Predictions (cont.)}
    \textbf{Variable selection}. We select \(p\) variables using only a point estimator of \(\tau|\mathcal D_n\) (so-called \textit{impact points}), and then we can apply any multiple regression algorithm to the reduced data set.
    \vspace{1em}

    \begin{figure}
      \includegraphics[width=\textwidth]{credible_intervals}
      \caption{Approximate posterior distribution for the impact points.}
    \end{figure}

@@

\begin{frame}{Affine-invariant ensemble sampler (emcee)}
Consider an \maroon{ensemble} \ of walkers \(\Lambda(t) = (\Lambda_1(t), \dots, \Lambda_L(t))\) and the complementary ensemble
\[
  \Lambda_{-l}(t) = \{\Lambda_1(t+1), \dots, \Lambda_{l-1}(t+1), \Lambda_{l+1}(t), \dots, \Lambda_L(t)\}, \quad l=1,\dots, L.
\]
\vspace{-1em}
\begin{enumerate}
  \item For each walker \(1\leq l \leq L\) another walker \(\Lambda_j \in \Lambda_{-l}(t)\) is chosen at random. The proposal is constructed as
  \[
    \Lambda_l(t) \to \Gamma = \Lambda_j + Z(\Lambda_l(t) - \Lambda_j),
\]
where \(Z \stackrel{i.i.d.}{\sim} g(z)\) satisfying \(g(z^{-1})=zg(z)\).
\item  The proposal is accepted with probability
  \[
    \alpha = \min\left\{1, \ Z^{p-1}\frac{\pi(\Gamma)}{\pi(\Lambda_l(t))}\right\}.
  \]
\end{enumerate}
\end{frame}

>>>>

\usepackage{fontspec}
\usepackage[autostyle=true]{csquotes}
\usepackage[absolute,overlay]{textpos}
\usepackage[english]{babel}
\usepackage{microtype}
\usepackage{subcaption}
\usepackage{epigraph}
\usepackage[export]{adjustbox}
\usepackage{hyperref}
\usepackage{bm}
\usepackage{amssymb, amsmath, amsthm, amsfonts, amscd, dsfont}

@@

\usetheme[block=fill, subsectionpage=progressbar, titleformat section=smallcaps]{metropolis}
\setbeamertemplate{frametitle continuation}[roman]
\setbeamerfont{section title}{size=\huge}
\setbeamertemplate{section in toc}[sections numbered]
%\setbeamertemplate{subsection in toc}[subsections unnumbered]
\widowpenalties 1 10000
\raggedbottom

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\renewcommand{\baselinestretch}{1.3}

% COLORS

\definecolor{Maroon}{cmyk}{0, 0.87, 0.88, 0.1}
\definecolor{teal}{rgb}{0.0, 0.45, 0.45}
\definecolor{mLightBrown}{HTML}{f97e0b}
% \definecolor{mLightBrown}{HTML}{990000}  % Darker accent color
% \definecolor{backg}{HTML}{F2F2F2} % Fondo
% \definecolor{comments}{HTML}{a8a8a8} % Comentarios
% \definecolor{keywords}{HTML}{08388c} % Palabras clave
% \definecolor{strings}{HTML}{0489B1}  % Strings
\setbeamercolor{palette primary}{bg=teal}
\setbeamercolor{progress bar}{use=Maroon, fg=Maroon}

% BIBLIOGRAPHY

@@

% GRAPHICS

@@

@@

\newcommand\maroon[1]{\color{mLightBrown}#1\color{mDarkTeal}}
\renewcommand{\epsilon}{\varepsilon}

@@

% DOCUMENT

@@

\begin{frame}{Functional linear regression}
  \maroon{First problem}: linear regression with functional data.
  \vspace{2em}

@@

\begin{frame}{Functional logistic regression}
  \maroon{Second problem}: logistic regression with functional data.
\vspace{2em}

@@

\begin{frame}{Shortcomings of the \(\bm{L^2}\)-models}
  \begin{itemize}
    \item Direct optimization is hopeless: \(L^2[0,1]\) is a \maroon{extremely wide space}.
    \item The usual least-squares theory cannot be applied directly; some regularization or dimensionality reduction techniques are required.
    \item They do not include simple models based on linear combinations of the marginal of the process as particular cases, such as
    \[
      Y = \alpha_0 + \sum_{j=1}^p \beta_j X(t_j) + \varepsilon.
    \]
    \item In the logistic case, the maximum likelihood estimator does not exist with probability 1 under fairly general conditions.
  \end{itemize}
\end{frame}

@@

\vspace{1em}
  We consider a functional parameter \maroon{\(\alpha \in \Hcal(K)\)}, where \(\Hcal(K)\) is the RKHS associated with the covariance function \(K(t, s)=\mathbb E[X(t)X(s)]\) of the centered process \(X\). Moreover, we substitute \(\dotprod{X}{\beta}_2\) for \maroon{\(\dotprod{X}{\alpha}_K\)}.
\end{frame}

@@

  \begin{block}{Proposition}
    \begin{enumerate}
      \item All the evaluation operators \(\delta_t(f)=f(t)\) are continuous \maroon{(characterization)}.
      \item Norm convergence implies pointwise convergence.
      \item If \(K\) is continuous, then so is every function \(f\in\Hcal(K)\).
    \end{enumerate}
  \end{block}
\end{frame}

@@

  Generally speaking, the concrete realizations \(x=X(\omega)\) \maroon{do not belong to \(\Hcal(K)\) with probability 1}, so the expression \(\dotprod{x}{\alpha}_K\) lacks meaning. However, we can make use of the relationship between \(\Hcal(K)\) and \(\Lcal(X)\), the linear span of the process \(X\) within \(L^2(\Omega)\).

  \vspace{1em}
  \begin{block}{Loève's isometry}
    It is the continuous extension of the correspondence
    \[
\sum_{j=1}^p a_j X(t_j) \longleftrightarrow \sum_{j=1}^p a_j K(t_j, \cdot).
    \]
    Formally, \(\Psi^{-1}_X(U)(t) := \E[U X(t)]\) for \(U \in \Lcal(X)\).
  \end{block}
  With a slight abuse of notation, we will write \maroon{\(\dotprod{x}{\alpha}_K \equiv [\Psi_X(\alpha)](\omega) := \Psi_x(\alpha)\)}\ for \(x=X(\omega)\) and \(\alpha \in \Hcal(K)\).
\end{frame}

@@

\begin{frame}{\ldots and two new models}
  Given a data set \(\D_n = \{(X_i, Y_i): i=1,\dots, n\}\) composed of i.i.d. samples of \((X, Y)\), for \(\alpha \in \Hcal_0(K)\) we consider the following models:

@@

      \begin{block}{Bayes's formula}
      \[
      \pi(\theta \mid \D_n) \propto \left( \prod_{i=1}^n \pi(Y_i\mid X_i, \theta) \right)\pi(\theta).
      \]
  \end{block}

@@

\begin{frame}{Prior distributions}
  Same prior scenario for both linear and logistic models.
  \vspace{.5em}
  \begin{block}{Prior for the parameters involved}
    \vspace{-1em}
  \begin{align*}
  \pi(\alpha_0, \sigma^2)              & \propto 1/\sigma^2,                                                     \\
  \tau                     & \sim \U([0, 1]^p),                                              \\
  b\mid \tau, \sigma^2 & \sim \mathcal N_p\big(b_0, g\sigma^2{\underbrace{\left[\X_\tau' \X_\tau + \eta I\right]}_{G_\tau}}^{-1}\big),
\end{align*}

\vspace{-1em}
where \(I\) is the identity matrix, \(\X_\tau = (X_i(t_j))_{i,j}\) is the data matrix, and \(b_0\in \R^p, \ g \in \R\) and \(\eta \in \R^+\) are hyperparameters.
\end{block}

@@

\begin{frame}{Posterior distributions}
  \begin{block}{Linear log-posterior}
    \vspace{-1em}
    \begin{align*}
  \log \pi(\theta\mid \D_n) & \propto \frac{1}{2}\log |G_\tau| - (p+n+2)\log \sigma\\
  &-\frac{1}{2\sigma^2} \left(\|\bm{Y}-\alpha_0\bm{1} - \X_\tau b\|^2 + \frac{1}{g}(b - b_0)'G_\tau(b - b_0) \right),
\end{align*}
where \(\bm Y=(Y_1,\dots,Y_n)'\) and \(\bm{1}\) is an \(n\)-dimensional vector of ones.
  \end{block}
  \vspace{1em}
  \begin{block}{Logistic log-posterior}
      \vspace{-1em}
    \begin{align*}
  \log \pi(\theta \mid \D_n) & \propto \sum_{i=1}^n \left[ \left(\alpha_0 + \dotprod{X_i}{\alpha}_K\right)Y_i - \log\left(1 + \exp\left\{\alpha_0 + \dotprod{X_i}{\alpha}_K\right\}\right)\right]\\
  \quad &+ \frac{1}{2}\log |G_\tau| - (p+2)\log \sigma -\frac{1}{2g\sigma^2} (b - b_0)'G_\tau(b - b_0).
\end{align*}
Remember that \(\dotprod{X_i}{\alpha}_K = \sum_j \beta_j X_i(t_j)\).
  \end{block}
\end{frame}

@@

  \begin{figure}
    \centering
    \includegraphics[width=.9\textwidth]{posterior}
  \end{figure}

@@

\begin{frame}{Predictions}
  \textbf{Summarize-then-predict}. A summary of the posterior distribution \(\theta| \D_n\) is obtained via a point-estimate statistic (mean, median, mode), and then predictions are made according to the assumed model, e.g.:
  \[
  \hat Y_i =\hat \alpha_0 + \sum_{j=1}^p \hat \beta_j X_i(\hat t_j), \quad i=1,\dots, n.
  \]

@@

\begin{frame}{Predictions (cont.)}
    We are essentially approximating the \textit{posterior predictive distribution}
\[
\pi(y\mid x, \D_n) = \int_{\Theta} \pi(y\mid x, \theta)\pi(\theta\mid \D_n)\, d\theta.
\]
  \begin{figure}
    \includegraphics[width=.675\textwidth]{ppc_linear}
  \end{figure}
\end{frame}

\begin{frame}{Predictions (cont.)}
    \textbf{Variable selection}. We select \(p\) variables using only a point estimator of \(\tau|\D_n\) (so-called \textit{impact points}), and then we can apply any multiple regression algorithm to the reduced data set.
    \vspace{1em}

    \begin{figure}
      \includegraphics[width=\textwidth]{credible_intervals}
    \end{figure}

@@

\begin{frame}{Affine-invariant ensemble sampler (emcee)}
Consider an \maroon{ensemble} \ of walkers \(\Lambda(t) = (\Lambda_1(t), \dots, \Lambda_L(t))\) and the complementary ensemble
\[
  \Lambda_{-l}(t) = \{\Lambda_1(t+1), \dots, \Lambda_{l-1}(t+1), \Lambda_{l+1}(t), \dots, \Lambda_L(t)\}, \quad l=1,\dots, L.
\]
\vspace{-1em}
\begin{enumerate}
  \item For each walker \(1\leq l \leq L\) another walker \(\Lambda_j \in \Lambda_{-l}(t)\) is chosen at random. The proposal is constructed as
  \[
    \Lambda_l(t) \to \Gamma = \Lambda_j + Z(\Lambda_l(t) - \Lambda_j),
\]
where \(Z \stackrel{i.i.d.}{\sim} g(z)\) satisfying \(g(z^{-1})=zg(z)\).

\item If \(\R^p\) is the sample space, the proposal is accepted with probability
  \[
    \alpha = \min\left\{1, \ Z^{p-1}\frac{\pi(\Gamma)}{\pi(\Lambda_l(t))}\right\}.
  \]
\end{enumerate}
\end{frame}
